Supplementary figures\nCapturing regeneration in models of forest dynamics

### 2023-04-25 manuscript/supplementaryRW.Rmd
corrections in 4C, authors names and others

---
title: "Supplementary figures\nCapturing regeneration in models of forest dynamics"
author: 
- Olalla Díaz-Yáñez, 
- Yannek Käber, 
- Mats Mahnken, 
- Samuel Fischer, 
- Xavier Morin, 
- Manfred J. Lexer, 
- Christian Hochauer, 
- Katarina Merganicova, 
- Tobias Mette, 
- Werner Rammer, 
- Kristin H. Braziunas, 
- Paola Mairota, 
- Josef Bruna,
- Heike Lischke, 
- Daniel Scherrer, 
- Simon Scheiter, 
- Thomas Hickler, 
- Tim Anders,
- Jessica Hetzer, 
- Harald Bugmann
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    latex_engine: xelatex
    extra_dependencies: ["flafter", "float", "subfig"]
    toc: yes
bibliography: Regeneration.bib
link-citations: yes
linkcolor: blue
csl: global-change-biology.csl
editor_options: 
  markdown: 
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
- \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

We used the provided daily ERA5-CHELSA climate data set and complemented
the CHELSA time series data regarding relative humidity with the model
after Eccel et al. (2015) based on minimum and mean daily temperature to
accomodate data needs of 4C. The missing data for January 2nd and 3rd
2013 were filled with data from January 1st and 4th 2013 respectively.
For construction of longer climate time series we looped through the
provided time series from 1981 to 2018 up to the maximum simulation
length of 2500 years. We amended situations where leap year
inconsistencies arised by removing February 29 or duplicating data from
February 28. Otherwise, the unaltered daily data were used to run 4C on
a daily basis. We applied a constant atmopsheric CO2 concentration of
380 ppm.

@@

We used parametrized soils readily available for 4C from the EU soil
data base. The link from parametrized soil to soil quality was done by
extracting the plant available water storage capacity as the model
bucket size of the rooted soil horizons. We excluded bucket size values
of lower than 10 cm and larger than 35 cm. The bucket size was
reprojected to the range from 0 to 1. In addition we tested the
sensitivity of dominant tree height at age 100 years for monospecific
stands of all five species descibed above to the available soils using
model simulations with 4C. The dominant tree height at age 100 years was
then also reprojected to the range from 0 to 1 in order to derive a
model specific soil quality indicator for all paramterized soils. Then
we multiplied this relative model specific soil quality with the
relative soil quality derived from the bucket size to the range from 1
to 5. The match between paramterized soil and plot was then done by
finding the pair of soil quality with the minimum difference between
simuation protocol prescribed soil quality and and the soil quality of
the paramterized soils as described above.

@@

There is no explicitly modelled seed dispersal in a stand and between
stands. We applied a constant sapling establishment rate every five
years of 2000 individuals/ha for all five species; 10000 individuals/ha
in total every five years. The constant sapling establishment provides
constant regeneration potential if favorable conditions for sapling
growth (especially light and water regime) are present. This approach
enables regeneration of other species than only the one present in the
stand to establish if competition allows. The resulting recruitment
patterns are an outcome of the dynamical processes from the sapling
stage onward. Seed production and germination were not explicitly
simulated.

@@

To reach equilibrium and obtain dynamics in the equilibrium period, we
simulated a maximum of 2500 years. 4C is usually applied for much
shorter simulation periods and different technical as well as
model-theory related issues arise when simulating longer than 100-150
years. Therefore 2500 years can be seen as the upper limit of continuous
simulation in the defined simulation setup. The first 500 years were not
sampled for recruitment as we assume the model needs this time to reach
equilibrium. In this long simulation period 4C partially projects
unrealistically large tree heights resulting in premature termination of
the simulation (36 plots) so that not all plots were simulated for the
full 2500 years. We started sampling the recruits after the first 500
simulation years from non-overlapping 10 year periods to obtain 200
samples. For shorter simulation periods we produced 200 samples from
overlapping 10 year periods. In addition, we excluded those plots in
which 4C simulated unrealistically high basal area values of over 150 m²
(19 plots). We used this large threshold to give the model maximum
flexibility to explore the solution space in these long simulation runs
although we acknowledge that this is an unrealistic overestimation that
is probably originating from misrepresented density dependent mortality
processes in long simulation runs and misrepresented single tree
dimensions due to wrong assumptions of tree geometry and biomass share
of such old-aged individuals. Furthermore those plots with less than
1000 simulation years (16 plots) before termination were excluded
because the sample space after the first 500 years was too small.
Additionally we included another subset criterion removing all samples
that have a ba \> 100 to exclude unrealistically high samples without
the need to exclude the full site. This resulted in discarding 0.8 % of
the sample

@@

We provide 200 samples each for 165 of the 200 plots (see reason for
excluded plots in "Simulation length and equilibrium criteria"). These
stem from 10 year periods that overlap partially (for \< 2500 simulation
years) or do not overlap (for 2500 simulation years). The number of
recruits (r.trees) is the sum of individuals that cross the threshold
diameter (7/10 cm) in the 10 year sample period per hectare that have
not already died until the sample year (Fig. 3). The recruit's basal
area (r.ba) is the sum of the basal area per hectare of the trees that
cross the diameter threshold that have not already died until the sample
year. The total stand basal area (ba) is the basal area per hectare of
all trees in the stand \> 0 cm diameter at breast height (no lower
boundary) at the sample time per species. We provide the data for both
diameter thresholds, 7 and 10 cm. In over 90 % of the decades no
recruitment takes place.

[4C
(v2.2)](%5Bhttps://doi.org/10.5194/gmd-13-5311-2020)](<https://doi.org/10.5194/gmd-13-5311-2020>))
is usually not used for simulations of forest equillibrium dynamics,
potential natural vegetation assessment and long simulation periods but
rather shorter-term simulations (up to 100-150 years). In addition, only
five species out of the pool of eleven potentially relevant species here
are parameterized in 4C so that projections of the potential natural
vegetation done wih 4C need to be considered in the light of model
uncertainty. For example, the spatial homogeneity in the model leads to
cyclic recruitment patterns (see Fig. 4). When the adult stands dies of,
young individuals are recruited into the larger diameter classes because
more light reaches lower canopy layer since larger (adult) trees do not
anymore prevent the light reaching lower canopy layers. This results in
increased growth of the regeneration layer during those times. In most
stands Fagus sylvatica dominates the stand after some time but
occasionally individuals of other species, like Picea abies or Quercus
spp. cross the recruitment diameter threshold.

@@

Please noticed that we included 0 values for species missing in a given
sample and for samples in sites where Landis did not simulate any tree.

@@

Jessica Hetzer, Thoms Hickler, Tim Anders

@@

We used the provided daily time series of climatic variables from 1981
to 2018. The climatic variables base on the CHELSA data set with a
spatial resolution of 0.008° x 0.008° (\~ 1 km) (Karger et al., 2021).
Daily average temperature (tas) [°C], daily average maximum temperature
(tasmax) [°C], daily average minimum temperature (tasmin) [°C],
precipitation sum (pr) [mm] and surface solar (shortwave) radiation
(rsds) were taken to set up the LPJ-GUESS simulation. Used climatic
variables were converted to netCDF files. As LPJ-GUESS requires the
climatic variables in specific units, temperature variables (tas,
tasmax, tasmin) [°C] were converted to Kelvin and surface solar
radiation (rsds) [J/m2] was converted to W/m2 by dividing rsds by 86400
s (3600 s/h \* 24 h). Nitrogen deposition was considered to be constant
at a level of 10 kgN/ha/year. Atmospheric carbon dioxide concentration
data from 1900 to 2018 was taken from the global carbon project (Quéré
et al. 2018).

@@

All sites We used were simulated with default values for a medium
textured soil type (sand: 0.35, clay: 0.15, silt: 0.5) in LPJ-GUESS.

@@

LPJ-GUESS can be run in different modes, with different levels of
abstraction of the population and community processes. For this study,
the more detailed 'cohort' mode was used, in which individuals, patches
and vertical canopy structure are represented explicitly, but living
individuals within a cohort (age class) of a given tree species in a
given patch are assumed to be identical (in terms of all state
variables, such as height and stem diameter). Simulations consider the
main tree species as well as other PFTs yielding an appropriate
representation of European vegetation, including only tree and shrub
species that are widely distributed across Europe and can become
dominant in some areas (Hickler et al., 2012). Mediterranean rain-green
small shrubs with shallow roots (e.g. Lavendula spp., Rosmarinus spp.)
and alpine/arctic shrubs (e.g. Vaccinium spp.) were represented as PFTs
in the model. As in earlier applications of LPJ-GUESS, herbaceous
vegetation was represented by two 'generic herb' PFTs, with C3 and C4
photosynthesis, respectively. The final set included 16 tree species,
one Mediterranean shrub PFT, one boreal/alpine shrub PFT, and the two
herbaceous PFTs described earlier (Table 2). To increase the chances of
shade-intolerant species becoming established, patch-destroying
disturbances were allowed as described in Hickler et al. 2012. Acer
pseudoplatanus and Alnus glutinosa are not parameterized in LPJ-GUESS
and therefore not considered.we did not simulate these species.

@@

Please describe how have you calculated each of the variables in the
output data.

The generated output file is called Output_SGN_LPJGUESS.txt. Each row of
the output dataset contains variables regarding the decade 2008-2018.
The first two columns describe the study site (column "site", values
reflect site ID) and the 1 ha samples of that specific site (column
"sample", values range from 1 to 200). Output variables are
species-specific. Although we simulated a total of 20 European tree
species and PFTs, our output file considers only the species required by
the protocol (column "species", values are the abbreviations for the
species name ("abal" for Abies alba, "cabe" for Carpinus betulus, "fasy"
for Fagus sylvatica, "frex" for Fraxinus excelsior, "pabi" for Picea
abies, "pisy" for Pinus sylvestris , "tico" for Tilia cordata ,
"betu_pen" for Betula pendula, "betu_pen" for Betula pubescens,
"quer_il" for Quercus ilex, "quer_pub" for Quercus pubescens, and
"quer_rob" for Quercus robur). Since Acer pseudoplatanus ("acps") and
Alnus glutinosa ("algl") are not parameterized for LPJ-GUESS so far, we
have not included these species in the output. Further we excluded
shrubs (alpine/arctic shrubs and Quercus coccifera) and herbaceous
vegetation from the final output as this study focusses on tree
regeneration only. The first variable is the number of 7 cm recruits
"r_7cm.trees" calculated as the sum of all trees of the same species
that cross the threshold of 7 cm DHB between 2008 and 2018 in that
specific sample of that specific site. The variable "r_10cm.trees"
refers to 10 cm threshold respectively. The 7 cm recruits' basal area
„r_7cm.ba" is calculated as a weighted sum of r_7cm.trees multiplied by
recruits basal area at that time (with ba=(dbh/2)∙π). Similarly,
r_10cm.ba was calculated for 10 cm recruits. The last two columns
describes the mean annual basal area and the mean number of trees per
site, sample, and species overall years between 2008 and 2018 (column
"ba").

Table 2: Species/PFT characteristics and parameter of Hickler et al.
2012. r base respiration rate, kallm1 constant in allometry equations,
Tc,min minimum winter temperature for establishment, Tc,max maximum
coldest-month temperature for establishment, GDD5 minimum degree-day sum
above 5 °C for establishment, fAWC minimum growing-season fraction of
available soil water holding capacity in the first soil layer, rfire
fraction of individuals surviving fire, kla:sa leaf longevity, CAmax
maximum woody crown area. Parameter that are associated with
establishment processes are marked in bold. Species/PFT

@@

```{r mortality3 , echo=FALSE, fig.cap="Recruitment ration between 7 and 10cm and the initial recruitment tree number at 7cm. Dashed red lines mark a ratio equal to 1 indicating no decrease between 7 and 10cm tree recruitment and ratio equal to 1.77 corresponding with the Reineke selfthinning ratio under evenaged conditions", out.width='70%', fig.pos='H'}
knitr::include_graphics(here::here("figures", "mort7_10_trends.png"))
```

>>>>

---
title: "Supplementary figures\nCapturing regeneration in models of forest dynamics"
author: 
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred J. Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Christopher P.O. Reyer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    latex_engine: xelatex
    extra_dependencies: ["flafter", "float", "subfig"]
    toc: yes
bibliography: Regeneration.bib
link-citations: yes
linkcolor: blue
csl: global-change-biology.csl
editor_options: 
  markdown: 
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
- \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

We used the provided daily ERA5-CHELSA climate data set and complemented
the CHELSA time series data regarding relative humidity with the model
after [Eccel et al. (2012)](https://doi.org/10.1002/met.258) based on
minimum and mean daily temperature to accomodate data needs of 4C. The
missing data for January 2nd and 3rd 2013 were filled with data from
January 1st and 4th 2013 respectively. For construction of longer
climate time series we looped through the provided time series from 1981
to 2018 up to the maximum simulation length of 2500 years. We amended
situations where leap year inconsistencies arised by removing February
29 or duplicating data from February 28. Otherwise, the unaltered daily
data were used to run 4C on a daily basis. We applied a constant
atmopsheric CO2 concentration of 380 ppm.

@@

We used parametrized soils readily available for 4C from the EU soil
data base. The link from parametrized soil to soil quality was done by
extracting the plant available water storage capacity as the model
bucket size of the rooted soil horizons. We excluded bucket size values
of lower than 10 cm and larger than 35 cm. The bucket size was
reprojected to the range from 0 to 1. In addition we tested the
sensitivity of dominant tree height at age 100 years for monospecific
stands of all five species simulated to the available soils using model
simulations with 4C. The dominant tree height at age 100 years was then
also reprojected to the range from 0 to 1 in order to derive a model
specific soil quality indicator for all paramterized soils. Then we
multiplied this relative model specific soil quality with the relative
soil quality derived from the bucket size to the range from 1 to 5. The
match between paramterized soil and plot was then done by finding the
pair of soil quality with the minimum difference between simuation
protocol prescribed soil quality and the soil quality of the
paramterized soils as described above.

@@

There is no explicitly modelled seed dispersal in a stand and between
stands. We applied a constant sapling establishment rate every five
years of 2000 individuals/ha for all five species; 10000 individuals/ha
in total every five years. The constant sapling establishment provides
constant regeneration potential if favorable conditions for sapling
growth (especially light and water regime) are present. This approach
enables regeneration of other species than only those present in the
stand to establish if competition allows. The resulting recruitment
patterns are an outcome of the dynamical processes from the sapling
stage onward. Seed production and germination were not explicitly
simulated.

@@

To reach equilibrium and obtain dynamics in the equilibrium period, we
simulated a maximum of 2500 years. 4C is usually applied for much
shorter simulation periods and different technical as well as
model-theory related issues arise when simulating longer than 100-150
years. Therefore 2500 years can be seen as the upper limit of continuous
simulation in the defined simulation setup. The first 500 years were not
sampled for recruitment as we assume the model needs this time to reach
equilibrium. In this long simulation period 4C partially projects
unrealistically large tree heights resulting in premature termination of
the simulation (36 plots) so that not all plots were simulated for the
full 2500 years. We started sampling the recruits after the first 500
simulation years from non-overlapping 10 year periods to obtain 200
samples. For shorter simulation periods we produced 200 samples from
overlapping 10 year periods. In addition, we excluded those plots in
which 4C simulated unrealistically high basal area values of over 150 m²
(19 plots). We used this large threshold to give the model maximum
flexibility to explore the solution space in these long simulation runs
although we acknowledge that this is an unrealistic overestimation that
is probably originating from misrepresented density dependent mortality
processes in long simulation runs and misrepresented single tree
dimensions due to wrong assumptions of tree geometry and biomass share
of such old-aged individuals. Furthermore those plots with less than
1000 simulation years (16 plots) before termination were excluded
because the sample space after the first 500 years was too small.
Additionally we included another subset criterion removing all samples
that have a ba \> 90 to exclude unrealistically high samples without the
need to exclude the full site. This resulted in discarding 0.8 % of the
samples.

@@

We provide 200 samples each for 48 plots and at least 190 samples each
for 117 plots while not simulating 35 plots (see reason for excluded
plots in \"Simulation length and equilibrium criteria\"). These stem
from 10 year periods that overlap partially (for \< 2500 simulation
years) or do not overlap (for 2500 simulation years). The number of
recruits (r.trees) is the sum of individuals that cross the threshold
diameter (7/10 cm) in the 10 year sample period per hectare that have
not already died until the sample year. The recruit's basal area (r.ba)
is the sum of the basal area per hectare of the trees that cross the
diameter threshold that have not already died until the sample year. The
total stand basal area (ba) is the basal area per hectare of all trees
in the stand \> 0 cm diameter at breast height (no lower boundary) at
the sample time per species. We provide the data for both diameter
thresholds, 7 and 10 cm. In over 90 % of the decades no recruitment
takes place.

[4C
(v2.2)](%5Bhttps://doi.org/10.5194/gmd-13-5311-2020)](<https://doi.org/10.5194/gmd-13-5311-2020>))
is usually not used for simulations of forest equillibrium dynamics,
potential natural vegetation assessment and long simulation periods but
rather shorter-term simulations (up to 100-150 years). In addition, only
five species out of the pool of eleven potentially relevant species here
are parameterized in 4C so that projections of the potential natural
vegetation done wih 4C need to be considered in the light of model
uncertainty. For example, the spatial homogeneity in the model leads to
cyclic recruitment patterns. When the adult stands dies of, young
individuals are recruited into the larger diameter classes because more
light reaches lower canopy layer since larger (adult) trees do not
anymore prevent the light reaching lower canopy layers. This results in
increased growth of the regeneration layer during those times. In most
stands Fagus sylvatica dominates the stand after some time but
occasionally individuals of other species, like Picea abies or Quercus
spp. cross the recruitment diameter threshold.

@@

Please notice that we included 0 values for species missing in a given
sample and for samples in sites where Landis did not simulate any tree.

@@

Tim Anders, Jessica Hetzer, Thomas Hickler

@@

We used the provided daily time series of climatic variables from 1981
to 2018. The climatic variables base on the CHELSA data set with a
spatial resolution of 0.008° x 0.008° (\~ 1 km) (Karger et al., 2021).
Daily average temperature (tas) [°C], daily average maximum temperature
(tasmax) [°C], daily average minimum temperature (tasmin) [°C],
precipitation sum (pr) [mm] and surface solar (shortwave) radiation
(rsds) were taken to set up the LPJ-GUESS simulation. Used climatic
variables were converted to netCDF files. As LPJ-GUESS requires the
climatic variables in specific units, temperature variables (tas,
tasmax, tasmin) [°C] were converted to Kelvin and surface solar
radiation (rsds) [Jm-2] was converted to W/m2 by dividing rsds by 86400
s (3600 s/h \* 24 h). Nitrogen deposition was considered to be constant
at a level of 10 kgN/ha/year. Atmospheric carbon dioxide concentration
data from 1900 to 2018 was taken from the global carbon project (Quéré
et al. 2018).

@@

All sites were simulated with default values for a medium textured soil
type (sand: 0.35, clay: 0.15, 823 silt: 0.5) in LPJ-GUESS.

@@

LPJ-GUESS can be run in different modes, with different levels of
abstraction of the population and community processes. For this study,
the more detailed 'cohort' mode was used, in which individuals, patches
and vertical canopy structure are represented explicitly, but living
individuals within a cohort (age class) of a given tree species in a
given patch are assumed to be identical (in terms of all state
variables, such as height and stem diameter). Simulations consider the
main tree species as well as other PFTs yielding an appropriate
representation of European vegetation, including only tree and shrub
species that are widely distributed across Europe and can become
dominant in some areas (Hickler et al., 2012). Mediterranean rain-green
small shrubs with shallow roots (e.g. *Lavendula* spp., *Rosmarinus*
spp.) and alpine/arctic shrubs (e.g. Vaccinium spp.) were represented as
PFTs in the model. As in earlier applications of LPJ-GUESS, herbaceous
vegetation was represented by two 'generic herb' PFTs, with C3 and C4
photosynthesis, respectively. The final set included 16 tree species,
one Mediterranean shrub PFT, one boreal/alpine shrub PFT, and the two
herbaceous PFTs described earlier (Table 2). To increase the chances of
shade-intolerant species becoming established, patch-destroying
disturbances were allowed as described in Hickler et al. 2012. *Acer
pseudoplatanus* and *Alnus glutinosa* are not parameterized in LPJ-GUESS
and therefore not considered.

@@

The generated output file is called Output_SGN_LPJGUESS.txt. Each row of
the output dataset contains variables regarding the decade 2008-2018.
The first two columns describe the study site (column "site", values
reflect site ID) and the 1 ha samples of that specific site (column
"sample", values range from 1 to 200). Output variables are
species-specific. Although we simulated a total of 20 European tree
species and PFTs, our output file considers only the species required by
the protocol (column "species", values are the abbreviations for the
species name ("abal" for Abies alba, "cabe" for Carpinus betulus, "fasy"
for Fagus sylvatica, "frex" for Fraxinus excelsior, "pabi" for Picea
abies, "pisy" for Pinus sylvestris , "tico" for Tilia cordata ,
"betu_pen" for Betula pendula, "betu_pen" for Betula pubescens,
"quer_il" for Quercus ilex, "quer_pub" for Quercus pubescens, and
"quer_rob" for Quercus robur). Since Acer pseudoplatanus ("acps") and
Alnus glutinosa ("algl") are not parameterized for LPJ-GUESS so far, we
have not included these species in the output. Further we excluded
shrubs (alpine/arctic shrubs and Quercus coccifera) and herbaceous
vegetation from the final output as this study focusses on tree
regeneration only. The first variable is the number of 7 cm recruits
"r_7cm.trees" calculated as the sum of all trees of the same species
that cross the threshold of 7 cm DBH between 2008 and 2018 in that
specific sample of that specific site. The variable "r_10cm.trees"
refers to 10 cm threshold respectively. The 7 cm recruits' basal area
„r_7cm.ba" is calculated as a weighted sum of r_7cm.trees multiplied by
recruits basal area at that time (with ba=(dbh/2)∙π). Similarly,
r_10cm.ba was calculated for 10 cm recruits. The last two columns
describes the mean annual basal area and the mean number of trees per
site, sample, and species overall years between 2008 and 2018 (column
"ba").

Table 1: Species/PFT characteristics and parameter of Hickler et al.
2012. r base respiration rate, kallm1 constant in allometry equations,
Tc,min minimum winter temperature for establishment, Tc,max maximum
coldest-month temperature for establishment, GDD5 minimum degree-day sum
above 5 °C for establishment, fAWC minimum growing-season fraction of
available soil water holding capacity in the first soil layer, rfire
fraction of individuals surviving fire, kla:sa leaf longevity, CAmax
maximum woody crown area. Parameter that are associated with
establishment processes are marked in bold. Species/PFT

@@

```{r mortality3 , echo=FALSE, fig.cap="Recruitment ratio between 7 and 10cm and the initial recruitment tree number at 7cm. Dashed red lines mark a ratio equal to 1 indicating no decrease between 7 and 10cm tree recruitment and ratio equal to 1.77 corresponding with the Reineke self-thinning ratio under evenaged conditions", out.width='70%', fig.pos='H'}
knitr::include_graphics(here::here("figures", "mort7_10_trends.png"))
```
